---
title: "Introducing_nhanesA"
author: "Christopher Endres"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introducing_nhanesA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Background
nhanesA was developed to enable fully customizable retrieval of data from the National Health and Nutrition Examination Survey (NHANES). The survey is conducted by the National Center for Health Statistics (NCHS), and data are publicly available at: http://www.cdc.gov/nchs/nhanes.htm. NHANES data are reported in well over one thousand peer-reviewed journal publications every year. 

## NHANES Data
Since 1999, the NHANES survey has been conducted continuously, and the surveys during that period are referred to as "continous NHANES" to distinguish from several prior surveys. Continuous NHANES surveys are grouped in two-year intervals, with the first interval being 1999-2000.

Most NHANES data are in the form of tables in SAS 'XPT' format. The survey is grouped into five data categories that are publicly available, as well as an additional category (Limited access data) that requires written justification and prior approval before access. Package nhanesA is intended mostly for use with the publicly available data, but some information pertaining to the limited access data can also be retrieved.

The five publicly available data categories are:
- Demographics  (DEMO)
- Dietary       (DIET)
- Examination   (EXAM)
- Laboratory    (LAB)
- Questionnaire (Q)
The abbreviated forms in parentheses may be substituted for the long form in nhanesA commands.

### List NHANES Tables
To quickly get familiar with NHANES data, it is helpful to display a listing of tables.
Use nhanesTables to get information on tables that are available for a given category for a given year.
```{r}
suppressWarnings(library(nhanesA))
nhanesTables('EXAM', 2005)
```

Note that the two-year survey intervals begin with the odd year. For convenience, only a single 4-digit year is entered such that ``` nhanesTables('EXAM', 2005) ``` and ``` nhanesTables('EXAM', 2006)``` yield identical output. In the preceding example we see that the Examination data for the 2005-2006 survey consists of thirteen tables.

### List Variables in an NHANES Table

After viewing the output, we decide we are interested in table 'BMX_D' that contains body measures data.
To better determine if that table is of interest, we can display detailed information on the table contents using nhanesTableVars.
```{r}
nhanesTableVars('EXAM', 'BMX_D')
```
We see that there are 27 columns in table BMX_D. The first column (SEQN) is the respondent sequence number and is included in every NHANES table. Effectively, SEQN is a subject identifier that is used to join information across tables.
We now import BMX\_D along with the demographics table DEMO\_D.
```{r}
bmx_d  <- nhanes('BMX_D')
demo_d <- nhanes('DEMO_D')
```
We then merge the tables and compute average values by gender for several variables: 
```{r}
bmx_demo <- merge(demo_d, bmx_d)
options(digits=4)
aggregate(cbind(bmxht,bmxwt, bmxleg, bmxcalf, bmxthicr)~riagendr, bmx_demo, mean)
```

### Translation of Coded Values

NHANES uses coded values for many fields. In the preceding example, gender is coded as 1 or 2. To determine what the values mean, we can list the code translations for the gender field RIAGENDR in table DEMO_D
```{r}
nhanesTranslate('DEMO_D', 'RIAGENDR')
```

If desired, we can use nhanesTranslate to apply the code translation to demo\_d directly by assigning data=demo\_d. 
```{r}
levels(as.factor(demo_d$riagendr))
demo_d <- nhanesTranslate('DEMO_D', 'RIAGENDR', data=demo_d)
levels(demo_d$riagendr)
bmx_demo <- merge(demo_d, bmx_d)
aggregate(cbind(bmxht,bmxwt, bmxleg, bmxcalf, bmxthicr)~riagendr, bmx_demo, mean)
```

### Downloading a Complete Survey
The primary goal of nhanesA is to enable fully customizable processing of select NHANES tables. However, it is quite easy to download entire surveys using nhanesA functions.
Say we want to download every questionnaire in the 2007-2008 survey.  We first get a list of the table names by using nhanesTables with namesonly = TRUE. The tables can then be downloaded using nhanes with lapply.
```{r,eval=FALSE}
q2007names  <- nhanesTables('Q', 2007, namesonly=TRUE)
q2007tables <- lapply(q2007names, nhanes)
names(q2007tables) <- q2007names
```

### Apply All Possible Code Translations to a Table
An NHANES table may have dozens of columns with coded values. Translating all possible columns is a three step process.
1: Download the table
2: Download the list of table variables using nhanesTableVars with namesonly=TRUE
3: Pass the table and variable list to nhanesTranslate

```{r}
bpx_d <- nhanes('BPX_D')
head(bpx_d[,6:11])
bpx_d_vars  <- nhanesTableVars('EXAM', 'BPX_D', namesonly=TRUE)
bpx_d <- suppressWarnings(nhanesTranslate('BPX_D', bpx_d_vars, data=bpx_d))
head(bpx_d[,6:11])
```

Some discretion is applied by nhanesTranslate such that not all of the coded columns will be translated. In general, columns that have at least two categories (e.g. Male, Female) will be translated.  In some cases the code translations are quite long, thus to improve readability the maximum translation string should be limited. The default translation string length is 32 but can be set as high as 128.  

#### Please send any feedback or requests to cjendres1@gmail.com. Hope you enjoy your experience with nhanesA!
Sincerely,  
Christopher Endres